Perceived value of video games, but not hours played, predicts mental well-being in adult Nintendo players

### 2024-07-10 ms.qmd
clean up for preprint

---
title: "Perceived value of video games, but not hours played, predicts mental well-being in adult Nintendo players"
shorttitle: "Nintendo gaming and well-being"
author:
  - name: Nick Ballou
    corresponding: false
    orcid: 0000-0003-4126-0696
    email: nick@nickballou.com
    affiliations:
      - name: University of Oxford
        department: Oxford Internet Institute
    roles:
      - conceptualization
      - data curation
      - methodology
      - formal analysis
      - writing

  - name: Matti Vuorre
    corresponding: false
    orcid: 0000-0001-5052-066X
    affiliations:
      - name: Tilburg University
      - name: University of Oxford
        department: Oxford Internet Institute
    roles:
      - methodology
      - funding acquisition
      - formal analysis
      - editing

  - name: Thomas Hakman
    corresponding: false
    orcid: 0009-0009-8292-2482
    email: nick@nickballou.com
    affiliations:
      - name: University of Oxford
        department: Oxford Internet Institute
    roles:
      - data curation
      - validation
      - editing

  - name: Kristoffer Magnusson
    corresponding: false
    orcid: 0000-0003-0713-0556
    affiliations:
      - name: Karolinska Institute
      - name: University of Oxford
        department: Oxford Internet Institute
    roles:
      - methodology
      - editing

  - name: Andrew K Przybylski
    corresponding: true
    email: andy.pryzbylski@oii.ox.ac.uk
    orcid: 0000-0003-4126-0696
    affiliations:
      - name: University of Oxford
        department: Oxford Internet Institute
    roles:
      - conceptualization
      - funding acquisition
      - project administration
      - editing
abstract: "Studies on video games and well-being often rely on self-report measures or data from a single game. Here, we study how 704 US adults' time spent playing over 150 1st-party Nintendo Switch games (totaling 140k hours) relates to their life satisfaction, affect, depressive symptoms, and general mental well-being. We replicate previous findings that playtime over the past 2 weeks does not predict well-being, and extend these findings to a wider range of timescales (1 hour to 1 year). Results suggest that relationships, if present, dissipate within 2 hours of gameplay. Our non-causal findings suggest substantial confounding would be needed to shift a meaningful true effect to the observed null. Although playtime was not related to well-being, players’ assessments of the value of game time—so called gaming life fit—was. Results emphasize the importance of defining the gaming population of interest, collecting data from more than one game, and focusing on how players integrate gaming into their lives rather than the amount of time spent."
keywords: [video games, digital trace data, well-being]
author-note:
  # Disclosures condensed to one paragraph, but you can start a field with two line breaks to break them up: \n\nNew Paragraph
  disclosures:
    study-registration: This study was registered on the Open Science Framework (https://osf.io/sjqyt).
    # Acknowledge and cite data/materials to be shared.
    data-sharing: Data and materials are available at XXXXX. 
    # Example: [Author name] has been a paid consultant for Corporation X, which funded this study.
    conflict-of-interest: Data for this study was provided by Nintendo of America. Nintendo of America had no role in the design of the study, the analysis of the data, or decision to publish any results. The authors declare no other competing interests. 
    # Example: This study was supported by Grant [Grant Number] from [Funding Source].
    financial-support: ~ This research was supported by Huo Family Foundation and the UK Economic and Social Research Council (ESRC) (ES/W012626/1). KM was supported by Forskningsrå det för hälsa, arbetsliv och välfärd (2021-01284).
    # Example: The authors are grateful to [Person] for [Reason].
    gratitude: ~
    # Example. Because the authors are equal contributors, order of authorship was determined by a fair coin toss.
    authorship-agreements: ~
bibliography: references.bib
floatsintext: true # If true, tables and figures are mingled with the text instead of listed at the end of the document.
numbered-lines: false # Numbered lines (.pdf and .docx only)
meta-analysis: false
format:
  apaquarto-docx: default
  # apaquarto-html:
  #   documentmode: man
  #   code-fold: true
  #   code-summary: "Show the code"
  #   wordcount-html: default
  # apaquarto-pdf:
  #   documentmode: jou
---

@@

datSurveyRaw <- read_csv(paste0(Sys.getenv("DATA_PATH"), "survey.csv.gz"))
datRaw <- read_csv(paste0(Sys.getenv("DATA_PATH"), "telemetry.csv.gz")) |> 
  left_join(datSurveyRaw, by = "pid")
datDemo <- read_csv(paste0(Sys.getenv("DATA_PATH"), "demographics.csv.gz"))

@@

@@

More specifically, our first hypothesis was to test whether the null relations reported in earlier work [@JohannesEtAl2021video; @VuorreEtAl2022Time; @BallouEtAl2024Registered] would replicate in an independent sample of play across the Nintendo platform. In our first hypothesis, we predicted that there will be no practically significant association between video game playtime in the previous 2 weeks and life satisfaction (H1a), affective valence (H1b), depressive symptoms (H1c), or general mental well-being (H1d).

@@

  ## Calculate well-being variables
  mutate(
    genderRC = ifelse(gender %in% c("Man","Woman",NA), gender, "Non-binary"),
    across(starts_with(c("wemwbs","promis","trojan","bangs","displacement")), ~case_when(
      . %in% c("Greatly interfered") ~ -3,
      . %in% c("Moderately interfered") ~ -2,
      . %in% c("Slightly interfered") ~ -1,
      . %in% c("No impact") ~ 0,
      . %in% c("1 - None of the time","Never","1 - Strongly disagree","1 \nStrongly Disagree","Slightly supported","1") ~ 1,
      . %in% c("2 - Rarely","Rarely","Moderately supported","2") ~ 2,
      . %in% c("3 - Some of the time","Sometimes","Greatly supported","3") ~ 3,
      . %in% c("4 - Often","Often","4Neither Agree nor Disagree","4") ~ 4,
      . %in% c("5 - All of the time","Always","5 - Strongly agree","5") ~ 5,
      . %in% c("6") ~ 6,
      . %in% c("7 Strongly agree") ~ 7,
      TRUE ~ NA_integer_))
  ) |> 
  rowwise() |> 
  mutate(
    meanWemwbs = mean(c_across(starts_with("wemwbs"))),
    meanPromis = mean(c_across(starts_with("promis"))),
    meanLifeFit = mean(c_across(starts_with("displacement")))
  ) |> 
  ungroup() |> 
  mutate(
    csas = rescale(csas, from = c(0,10), to = c(1,5)),
    affectiveValence = rescale(affectiveValence_1, from = c(1, 100), to = c(1, 5)),
    .keep = "unused"
    ) %>%

@@

```{r}
#| label: fig-sankey
#| fig-align: center
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "Participant flow, from recruitment to final sample"
#| apa-twocolumn: true

@@

**Affective valence** was measured with a single item: "How are you feeling right now?" [@KillingsworthGilbert2010Wandering]. Participants responded using a 100-point visual analogue scale (VAS) with endpoints "Very bad" and "Very good", which we rescaled to 1-5 to match the other well-being measures.

@@

```{r}
#| label: fig-playtime
#| fig-cap: "Description of playtime in the sample. Panel A shows the distribution of playtime across players for each observation period, with the mean shown in red; observations periods below the dashed line are shown as total hours, whereas periods above the dashed line reflect hours of player per day. Panel B shows the proportion of players who logged at least one play session in that period."
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| fig-pos: "t"
#| apa-twocolumn: true

@@

```{r}
#| label: fig-scatter
#| fig-cap: Scatterplots depicting the relationship between video game playtime during the previous 2 weeks (mean minutes of play per day) and four types of well-being.
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| fig-pos: "t"
#| apa-twocolumn: true

@@

```{r}
#| label: fig-ests
#| fig-cap: Estimates for the relationship between playtime and well-being across various timescales, shown with 90% (dark blue) and 99% (light blue) confidence intervals. Dashed lines represent the positive and negative smallest effect size of interest (SESOI) of .06. 
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| fig-pos: "t"
#| apa-twocolumn: true

@@

```{r}
#| label: fig-lifefit
#| fig-cap: Marginal relationship between gaming life fit (perceived harmful or beneficial effects of games for oneself) and wellbeing 
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| fig-pos: "t"
#| apa-twocolumn: true

@@

preds <- rbind(
  plot_predictions(lifeFitCsas, condition = "meanLifeFit", draw = FALSE) |> mutate(model = "Life Satisfaction") |> rename(wb = csas),
  plot_predictions(lifeFitAffect, condition = "meanLifeFit", draw = FALSE) |> mutate(model = "Affective Valence") |> rename(wb = affectiveValence),
  plot_predictions(lifeFitPromis, condition = "meanLifeFit", draw = FALSE) |> mutate(model = "Depressive Symptoms") |> rename(wb = meanPromis),
  plot_predictions(lifeFitWemwbs, condition = "meanLifeFit", draw = FALSE) |> mutate(model = "General Mental Well-being") |> rename(wb = meanWemwbs)
)

ggplot(preds, aes(x = meanLifeFit, y = estimate)) +
  geom_point(
    data = dat |> pivot_longer(cols = c(affectiveValence, meanPromis, meanWemwbs, csas), 
                               names_to = "model", values_to = "wellbeing") |> 
      mutate(model = recode(model, 
                                "affectiveValence" = "Affective Valence",
                                "csas" = "Life Satisfaction",
                                "meanPromis" = "Depressive Symptoms",
                                "meanWemwbs" = "General Mental Well-being")), 
    aes(x = meanLifeFit, y = wellbeing, group = model), alpha = .1) +
  geom_line(color = "darkgreen", linewidth = 1.5) +
  geom_ribbon(
    aes(ymin = conf.low, ymax = conf.high),
    group = 1,
    alpha = .2,
    # fill = "darkred"
  ) +
  facet_wrap(~ model, scales = "free_y") +
  labs(x = "Mean Gaming Life Fit", y = "Wellbeing (1-5 Scale)") +
  mytheme

@@

We performed various sensitivity checks to ensure the robustness of our findings (see supplementary materials for full detail). First, we explored potential non-linearity in the relationships between playtime and mental health by comparing a generalized additive model of well-being with and without a smooth term for playtime, and comparing AIC between these. Of the 48 possible models (4 well-being variables \* 12 playtime windows), just 1 of these (playtime in the previous 1 year and life satisfaction) showed a difference in AIC of more than 2, indicating that nearly all relationships were adequately captured by linear terms. Next, we reran the analyses using session durations as calculated by Nintendo, as opposed to the implied duration based on the start and end timestamps; Nintendo's durations differ from the implied duration in approximately 10% of sessions, reflecting certain unknown heuristics used by Nintendo to calculate playtime (based e.g., on idle time, time spent in menus, and so on). Data show a similar pattern: no models showed a significant relationship between playtime and well-being at our specified alpha of .01. Next, we explored alternative models wherein playtime was separated into both a binary variable (1 if the player had any time logged in that period, 0 if not), and a continuous variable (how much a person played). Results were comparable; although three models indicated that among those who played in the previous 1-2 weeks, longer play is associated with higher affect and general mental well-being, the remaining 93 playtime variables were neither significant or nor within both equivalence bounds.

@@

metadata <- read_csv(paste0(Sys.getenv("DATA_PATH"), "gameMetadata.csv.gz"))

@@

>>>>

---
title: "Perceived value of video games, but not hours played, predicts mental well-being in adult Nintendo players"
shorttitle: "Nintendo gaming and well-being"
author:
  - name: Nick Ballou
    corresponding: false
    orcid: 0000-0003-4126-0696
    email: nick@nickballou.com
    affiliations:
      - name: University of Oxford
        department: Oxford Internet Institute
    # roles:
    #   - conceptualization
    #   - data curation
    #   - methodology
    #   - formal analysis
    #   - writing

  - name: Matti Vuorre
    corresponding: false
    orcid: 0000-0001-5052-066X
    affiliations:
      - name: Tilburg University
      - name: University of Oxford
        department: Oxford Internet Institute
    # roles:
    #   - methodology
    #   - funding acquisition
    #   - formal analysis
    #   - editing

  - name: Thomas Hakman
    corresponding: false
    orcid: 0009-0009-8292-2482
    email: nick@nickballou.com
    affiliations:
      - name: University of Oxford
        department: Oxford Internet Institute
    # roles:
    #   - data curation
    #   - validation
    #   - editing

  - name: Kristoffer Magnusson
    corresponding: false
    orcid: 0000-0003-0713-0556
    affiliations:
      - name: Karolinska Institute
      - name: University of Oxford
        department: Oxford Internet Institute
    # roles:
    #   - methodology
    #   - editing

  - name: Andrew K Przybylski
    corresponding: true
    email: andy.pryzbylski@oii.ox.ac.uk
    orcid: 0000-0003-4126-0696
    affiliations:
      - name: University of Oxford
        department: Oxford Internet Institute
    # roles:
    #   - conceptualization
    #   - funding acquisition
    #   - project administration
    #   - editing
abstract: "Studies on video games and well-being often rely on self-report measures or data from a single game. Here, we study how 704 US adults' time spent playing over 150 1st-party Nintendo Switch games (totaling 140k hours) relates to their life satisfaction, affect, depressive symptoms, and general mental well-being. We replicate previous findings that playtime over the past 2 weeks does not predict well-being, and extend these findings to a wider range of timescales (1 hour to 1 year). Results suggest that relationships, if present, dissipate within 2 hours of gameplay. Our non-causal findings suggest substantial confounding would be needed to shift a meaningful true effect to the observed null. Although playtime was not related to well-being, players’ assessments of the value of game time—so called gaming life fit—was. Results emphasize the importance of defining the gaming population of interest, collecting data from more than one game, and focusing on how players integrate gaming into their lives rather than the amount of time spent."
keywords: [video games, digital trace data, well-being]
author-note:
  # Disclosures condensed to one paragraph, but you can start a field with two line breaks to break them up: \n\nNew Paragraph
  disclosures:
    study-registration: This study was registered on the Open Science Framework (https://osf.io/sjqyt).
    # Acknowledge and cite data/materials to be shared.
    data-sharing: Data and materials are available at https://osf.io/6xkdg/. 
    # Example: [Author name] has been a paid consultant for Corporation X, which funded this study.
    conflict-of-interest: Data for this study was provided by Nintendo of America. Nintendo of America had no role in the design of the study, the analysis of the data, or decision to publish any results. The authors declare no other competing interests. 
    # Example: This study was supported by Grant [Grant Number] from [Funding Source].
    financial-support: This research was supported by Huo Family Foundation and the UK Economic and Social Research Council (ES/W012626/1). KM was supported by Forskningsrå det för hälsa, arbetsliv och välfärd (2021-01284).
    # Example: The authors are grateful to [Person] for [Reason].
    gratitude: ~
    # Example. Because the authors are equal contributors, order of authorship was determined by a fair coin toss.
    authorship-agreements: ~
bibliography: references.bib
floatsintext: true # If true, tables and figures are mingled with the text instead of listed at the end of the document.
numbered-lines: false # Numbered lines (.pdf and .docx only)
meta-analysis: false
suppress-credit-statement: true
format:
  apaquarto-docx: default
    # blank-lines-above-author-note: 0
    # blank-lines-above-title: 1
  # apaquarto-html:
  #   documentmode: man
  #   code-fold: true
  #   code-summary: "Show the code"
  #   wordcount-html: default
  # apaquarto-pdf:
  #   documentmode: jou
---

@@

datSurveyRaw <- read_csv("data/survey.csv.gz")
datRaw <- read_csv("data/telemetry.csv.gz") |> 
  left_join(datSurveyRaw, by = "pid")
datDemo <- read_csv("data/demographics.csv.gz")

@@

```{r}
#| label: fig-sankey
#| fig-align: center
#| out-width: "80%"
#| fig-cap: "Participant flow, from recruitment to final sample"
#| apa-twocolumn: true

@@

More specifically, our first hypothesis was to test whether the null relations reported in earlier work [@JohannesEtAl2021video; @VuorreEtAl2022Time; @BallouEtAl2024Registered] would replicate in an independent sample of Nintendo Switch play. In our first hypothesis, we predicted that there will be no practically significant association between video game playtime in the previous 2 weeks and life satisfaction (H1a), affect (H1b), depressive symptoms (H1c), or general mental well-being (H1d).

@@

  mutate(genderRC = ifelse(gender %in% c("Man","Woman",NA), gender, "Non-binary")) |> 
  rowwise() |> 
  mutate(
    meanWemwbs = mean(c_across(starts_with("wemwbs"))),
    meanPromis = mean(c_across(starts_with("promis"))),
    meanLifeFit = mean(c_across(starts_with("displacement"))),
    meanBPS = mean(c_across(starts_with("bps"))),
  ) |> 
  ungroup() |> 
  mutate(
    csas = rescale(csas, from = c(0,10), to = c(1,5)),
    affectiveValence = rescale(affectiveValence_1, from = c(1, 100), to = c(1, 5)),
    .keep = "unused"
    ) %>%

@@

@@

**Affect** was measured with a single item: "How are you feeling right now?" [@KillingsworthGilbert2010Wandering]. Participants responded using a 100-point visual analogue scale (VAS) with endpoints "Very bad" and "Very good", which we rescaled to 1-5 to match the other well-being measures.

@@

```{r}
#| label: fig-playtime
#| fig-cap: "Description of playtime in the sample. Panel A shows the distribution of playtime across players for each observation period, with the mean shown in red; observations periods below the dashed line are shown as total hours, whereas periods above the dashed line reflect hours of player per day. Panel B shows the proportion of players who logged at least one play session in that period."
#| fig-width: 8
#| fig-height: 4
#| fig-align: center
#| fig-pos: "t"
#| apa-twocolumn: true

@@

```{r}
#| label: fig-scatter
#| fig-cap: Scatterplots depicting the relationship between video game playtime during the previous 2 weeks (mean minutes of play per day) and four types of well-being.
#| fig-width: 6
#| fig-height: 4.5
#| fig-align: center
#| fig-pos: "t"
#| apa-twocolumn: true

@@

```{r}
#| label: fig-ests
#| fig-cap: Estimates for the relationship between playtime and well-being across various timescales, shown with 90% (dark blue) and 99% (light blue) confidence intervals. Dashed lines represent the positive and negative smallest effect size of interest (SESOI) of .06. 
#| fig-width: 6
#| fig-height: 4.5
#| fig-align: center
#| fig-pos: "t"
#| apa-twocolumn: true

@@

```{r}
#| label: fig-lifefit
#| fig-cap: Marginal relationship between gaming life fit (perceived harmful or beneficial effects of games for oneself) and wellbeing 
#| fig-width: 6
#| fig-height: 4.5
#| fig-align: center
#| fig-pos: "t"
#| apa-twocolumn: true

@@

preds <- rbind(
  plot_predictions(lifeFitCsas, condition = "meanLifeFit", draw = FALSE) |> mutate(model = "Life Satisfaction") |> rename(wb = csas),
  plot_predictions(lifeFitAffect, condition = "meanLifeFit", draw = FALSE) |> mutate(model = "Affective") |> rename(wb = affectiveValence),
  plot_predictions(lifeFitPromis, condition = "meanLifeFit", draw = FALSE) |> mutate(model = "Depressive Symptoms") |> rename(wb = meanPromis),
  plot_predictions(lifeFitWemwbs, condition = "meanLifeFit", draw = FALSE) |> mutate(model = "General Mental Well-being") |> rename(wb = meanWemwbs)
)

ggplot(preds, aes(x = meanLifeFit, y = estimate)) +
  geom_point(
    data = dat |> pivot_longer(cols = c(affectiveValence, meanPromis, meanWemwbs, csas), 
                               names_to = "model", values_to = "wellbeing") |> 
      mutate(model = recode(model, 
                                "affectiveValence" = "Affective",
                                "csas" = "Life Satisfaction",
                                "meanPromis" = "Depressive Symptoms",
                                "meanWemwbs" = "General Mental Well-being")), 
    aes(x = meanLifeFit, y = wellbeing, group = model), alpha = .1) +
  geom_line(color = "darkgreen", linewidth = 1.5) +
  geom_ribbon(
    aes(ymin = conf.low, ymax = conf.high),
    group = 1,
    alpha = .2,
    # fill = "darkred"
  ) +
  facet_wrap(~ model, scales = "free_y") +
  labs(x = "Mean Gaming Life Fit", y = "Wellbeing (1-5 Scale)") +
  mytheme

@@

We performed various sensitivity checks to ensure the robustness of our findings (see supplementary materials for full detail). First, we explored potential non-linearity in the relationships between playtime and mental health by comparing a generalized additive model of well-being with and without a smooth term for playtime, and comparing AIC between these. Of the 48 possible models (4 well-being variables \* 12 playtime windows), just 1 of these (playtime in the previous 1 year and life satisfaction) showed a difference in AIC of more than 2, indicating that nearly all relationships were adequately captured by linear terms. Next, we reran the analyses using session durations as calculated by Nintendo, as opposed to the implied duration based on the start and end timestamps; Nintendo’s durations are shorter than the implied duration in approximately 10% of sessions due to internal methodology. Data show a similar pattern: no models showed a significant relationship between playtime and well-being at our specified alpha of .01. Next, we explored alternative models wherein playtime was separated into both a binary variable (1 if the player had any time logged in that period, 0 if not), and a continuous variable (how much a person played). Results were comparable; although three models indicated that among those who played in the previous 1-2 weeks, longer play is associated with higher affect and general mental well-being, the remaining 93 playtime variables were neither significant or nor within both equivalence bounds.

@@

metadata <- read_csv("data/gameMetadata.csv.gz")

@@

# Contributions

Author roles were classified using the Contributor Role Taxonomy (CRediT;https://credit.niso.org/) as follows: Nick Ballou: conceptualization, data curation, methodology, formal analysis, writing; Matti Vuorre: methodology, funding acquisition, formal analysis, editing; Thomas Hakman: data curation, validation, editing; Kristoffer Magnusson: methodology, editing; Andrew K Przybylski: conceptualization,funding acquisition, project administration, editing
